---
title: "Statistical test for the heart failure dataset (Ahmad et al. 2017, PlosOne)"
author: "Giuseppe Jurman"
date: "March, 1st 2019"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

We perform here a quick statistical exploration of the heart failure dataset originally analysed in [Ahmad et al., PlosONE, 2017](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0181001)

To this aim we use an instance of the original data **without the time feature** and **not normalized**.

First we load data (*Davide, just set your correct path if you want to run the notebook*)
```{r, message=FALSE}
library(readr)
path_to_data <- "~/Google Drive/chicco_survival/data/"
filename <- paste(path_to_data,"dataset_edited_without_time.csv",sep="")
dataset_edited_without_time <- as.data.frame(read_csv(filename,))
print(head(dataset_edited_without_time))
```

Now we set *death_event* as a **factor**, since the statistical tests want a factor as the second element of the "~" formula.
```{r}
dataset_edited_without_time$death_event <- 
  as.factor(dataset_edited_without_time$death_event)
```

We proceed now with the computation of the four following statistical comparative tests:

* *Wilcoxon rank test* (aka Mann–Whitney U test), applied the feature *feat* versus *death_event* to detect whether we can reject the null hypothesis that the distribution of the feature *feat* for the groups of samples tdefined by *death_event* are the same; here we cannot use the *Wilcoxon signed rank test*, which needs the samples to be paired. To apply this last method, we should use repeated subsampling.
* *Kruskal Wallis test*, applied the feature *feat* versus *death_event* to detect whether we can reject the null hypothesis that the features *feat* of the samples grouped accordingly to *death_event* were selected from populations having the same distribution (it is indeed an extension of the Wilcoxon rank test);
* *Chi squared test*, applied the feature *feat* versus *death_event* to detect whether there is a significant association between the two variables; **disclaimer**: it should be applied to contingency vectors (counts), thus applying to numerical variables may by inappropriate. Note that we need to set the parameter *simulate.p.value = TRUE* to avoid warnings due to the small sample size.

plus the additional *Shapiro* test, which is applied to a *single* feature to detect whether *feat* has been extracted from a normal distribution.

All the outputs of the tests are stored on the *alltests* lists, that we print at the end of the discussion.

```{r}
mycols <- names(dataset_edited_without_time)
mycols <- mycols[mycols!="death_event"]

alltests <- list()

alltests[["Wilcoxon_rank"]] <- 
  alltests[["Kruskal"]] <- 
  alltests[["Chi"]] <- 
  alltests[["Shapiro"]] <- list()

for(thecol in mycols){
  alltests[["Wilcoxon_rank"]][[thecol]] <- 
    wilcox.test(as.formula(paste(thecol,"death_event",sep="~")), 
                data=dataset_edited_without_time)
  alltests[["Kruskal"]][[thecol]] <- 
    kruskal.test(as.formula(paste(thecol,"death_event",sep="~")), 
                 data=dataset_edited_without_time)
  alltests[["Chi"]][[thecol]] <- 
    chisq.test(x=as.factor(dataset_edited_without_time[,thecol]),
               y=dataset_edited_without_time$death_event,
               simulate.p.value = TRUE)
  alltests[["Shapiro"]][[thecol]] <- 
    shapiro.test(dataset_edited_without_time[,thecol])
}
alltests[["Shapiro"]][["death_event"]]<- 
  shapiro.test(as.numeric(dataset_edited_without_time$death_event))
```

### Discussion
As a rule of thumb, the validity of the tests is assessed by looking at the resulting $p$-values.

We start with the Shapiro test of normality, 
```{r}
dummy <- c()
for(thecol in c(mycols,"death_event")) dummy[thecol] <- alltests[["Shapiro"]][[thecol]]$p.value
print(dummy)
```
The very $p$-values indicate that all the variables of the dataset can be considered as extracted from a normal distribution.

We move now to the comparison of the *death_event* target with the other features of the dataset.

For the Wilcoxon rank test (that is, the Mann–Whitney U test) $p$-values,
```{r}
dummy <- c()
for(thecol in mycols) dummy[thecol] <-  round(alltests[["Wilcoxon_rank"]][[thecol]]$p.value,6)
print(dummy)
```
we can say that the values of each of the features *age*, *ejection_fraction*, *serum_sodium* and *serum_creatinine* are extracted from different distributions between the samples in groups *death_event=0* and *death_event=1*, while null hypothesis of same distribution for the two groups of samples cannot be ruled out for *gender*, *smoking*, *diabetes*, *blood-pressure*, *anaemia*, *platelets* and *creatinine_phosphokinase*.

For the Kruskal Wallis $p$-values,
```{r}
dummy <- c()
for(thecol in mycols) dummy[thecol] <- round(alltests[["Kruskal"]][[thecol]]$p.value,6)
print(dummy)
```
we can say that for *age*, *ejection_fraction*, *serum_sodium*, *serum_creatinine* the null hypothesis of same distribution for the two groups of samples *death_event=0* and *death_event=1* can be ruled out, while the same cannot be stated for *gender*, *smoking*, *diabetes*, *blood-pressure*, *anaemia*, *platelets* and *creatinine_phosphokinase*. This is consistent with the result of the Wilcoxon rank test, as expected.

For the $\chi^2$ $p$-values,
```{r}
dummy <- c()
for(thecol in mycols) dummy[thecol] <- round(alltests[["Chi"]][[thecol]]$p.value,6)
print(dummy)
```
again we have the same results of the Wilcoxon rank and Kruskal Wallis tests, that is, only *age*, *ejection_fraction*, *serum_sodium*, *serum_creatinine* have a significant relation with *death_event*, while for the other features the null hypothesis of independence cannot be discarded.

**Overall, we can conclude that these statistical tests indicate *age*, *ejection_fraction*, *serum_sodium*, *serum_creatinine* as the four important features of the dataset for discriminating survival as expressed by the *death_event* variable.**

For completeness, we print out all the tests' outputs.
```{r}
print(alltests)
```
